--- 
title: "Artifex - Fall 2023"
author: "David Reynolds"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
# output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: openscapes/series
description: "(Partial) course materials for Artifex, Spring 2023."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(tidyverse)
library(quantmod)
library(broom) # for the 'tidy' function
```


# Linear Regression Revisited

Let's investigate the relationship between home prices and mortgage rates. To start, let's use data from the <code>quantmod</code> package and look at the relationship between these two variables from 2000 to 2004.


```{r echo=TRUE}
# https://fred.stlouisfed.org/
pr = getSymbols('MORTGAGE30US',src='FRED', warnings = F) # 30-Year Fixed Rate Mortgage Average
home = getSymbols('CSUSHPINSA', src = 'FRED') # S&P/Case-Shiller U.S. National Home Price Index


# clean up dataset
mort = tidy(MORTGAGE30US) %>% # tidy is from the broom pkg - coerces xts object to a data frame
  mutate(date = ymd(index)) %>% # coerce the index to be a date
  filter(year(date) >= 2000, year(date) <= 2004) %>% # date range
  mutate(mo_year = ymd ( str_c (year(date), "-", month(date), "-", 01 )) ) %>% # make a new date so we can group by month/year 
  group_by(mo_year) %>% 
  summarise(val = mean(value)) %>% 
  filter(mo_year < "2023-08-01")


home_pr = tidy(CSUSHPINSA) %>% 
  mutate(date = ymd(index)) %>% 
  filter(year(date) >= 2000, year(date) <= 2004)

combined_dat = data.frame(mort = mort$val, pr = home_pr$value, date = home_pr$date)

ggplot(combined_dat, aes( mort, pr, color = as.factor( year(date) ) ) ) + 
  geom_point() + 
  theme_minimal() + 
  ggtitle("Case Shiller Index (y) and Avg. Mortgage Rates (x)") + 
  xlab("30-Year Fixed Rate Mortgage Average") + 
  ylab("S&P/Case-Shiller U.S. National Home Price Index")
```

How would you describe this relationship? Based on the relationship, do you think a simple linear regression model is appropriate for these data?

## Assignment 2

1. Fit a linear regression model to these two variables and write out the model equation.

2. Make a histogram of the residuals from this model and comment on how they correspond with the assumption of normal residuals.

3. Plot the residuals over time. What do you see?

4. Based on the residuals over time (plot from question 3), is there any additional variable you could add to the model to improve fit? If so, add this variable as a covariate to the linear regression and compare the fit of the augmented model with the one you fit in question 1.

5. Currently, the 30-Year Fixed Rate Mortgage Average is around 8%. What does the model predict for the Case-Shiller Index? The current value of the Case-Shiller Index is 305. What is the residual?

6. Now, using a scatterplot, check the relationship between these two variables for a more recent time frame (such as the last 10 years). Do you see the same relationship? Color code observations by year. 



# Kayak Data Analysis

We are going to do some analysis on a dataset sent from Kayak. The prompts and data are from their data science recruiting process. To access the data, use the following code:

```{r echo=TRUE, eval=FALSE}
visits = read.csv("https://raw.githubusercontent.com/dbreynol/DS809/main/data/visits.csv")
conversions = read.csv("https://raw.githubusercontent.com/dbreynol/DS809/main/data/conversions.csv")
channels = read.csv("https://raw.githubusercontent.com/dbreynol/DS809/main/data/channel_descriptions.csv")
```

## Prompt 

One of our team's goals is to accurately evaluate how successful our marketing campaigns are and to predict how successful future campaigns will be. In this assignment, we want you to explore how to evaluate one notion of success: conversion rate. Depending on the context, a conversion can mean different things, including a flight or hotel booking, a click on a specific part of our site, or an account sign up. Conversion rate in the first case could therefore be defined as completed flight bookings per user visit.
You were given some datasets that contain a sanitized version of some of our logging data. Explore the datasets and work on the questions below. One of the CSVs contains data on daily user visits broken down by user location and marketing channel, and another contains data on conversions also broken down by user location and channel. You also have been given a file with a brief description of the channels.

1. Do you see any interesting patterns in the data? For example, are there any seasonality trends in user visits or conversion rates (conversions/visits)? Do these seasonality trends differ between countries or channel? Can you think of any potential explanations for any of the patterns/trends/differences you saw?

2. Which channels have the best and worst conversion rates? What reasons can you think of to explain the differences? Do you see any major conversion differences for the same channel in different countries?

3. Predict next month's (September 2015) aggregate conversion rates for each channel and country combination? How good are your predictions?


```{r echo=F}
visits = read.csv("data/visits.csv")
conversions = read.csv("data/conversions.csv")
channels = read.csv("data/channel_descriptions.csv")
```



