# Set 1

idea- make a lab question about smoothing

## Time Series Data

* A time series is an ordered sequence of observations, where the ordering is through time.

* Time series data creates unique problems for statistical modeling and inference.
  + Traditional inference assumes that observations (data) are independent and identically distributed. Adjacent data points in time series data are not necessarily independent (uncorrelated).
  + Most time series models aim to exploit such dependence. For instance, yesterday’s demand of a product may tell us something about today’s
demand of a product.

* There are several different ways to represent time series data in R. 

* We will use the <code>tidyverse</code> family of packages extensively in this class. This package includes the <code>lubridate</code> package, which includes functions to work with date-times.

* Two of the most common ways to represent time series data are using data frames in which one of the variables is a time object (such as POSIXct or Date) or using a time series object. These two representations are shown below with simulated trading data for a single 8-hour trading day.


```{r echo=TRUE}
set.seed(1)

# option 1: represent time series data within a data frame

hr = seq(mdy_hm("12-11-2023 09:30"), mdy_hm("12-11-2023 16:30"), 'hour') # 8 hours
pr = rnorm(8) # generate fake trading data

trading_dat = data.frame(hr, pr)

# option 2: represent time series data using a time series object

trading_ts = ts(data = trading_dat$pr, start = 1, frequency = 8)
```



## Time Series EDA

The first thing to do in any data analysis is exploratory data analysis (EDA). Graphs enable many features of the data to be visualized, including patterns, unusual observations, changes over time, and relationships between variables. The features that are seen in plots of the data can then be incorporated into statistical models.

R has several systems for making graphs. We will primarily use ggplot2, which is among the set of tidyverse packages and is one of the most versatile systems for plotting. We will use a data set from Kayak to motivate our analysis.

```{r echo=TRUE}
conversions = read.csv("https://raw.githubusercontent.com/dbreynol/DS809/main/data/conversions.csv")
knitr::kable(head(conversions))
```

This dataset contains information on the total number of daily conversions by country and marketing channel. Let us focus our analysis on the US and fist visualize the number of conversions by day. 

```{r}
conversions$datestamp = ymd(conversions$datestamp)

# overall trend

conversions %>% 
  filter(country_code == "us") %>% 
  group_by(datestamp, country_code) %>% 
  summarise(tot_conv = sum(conversions), .groups = "drop") %>% 
  ggplot(aes(x = datestamp, y = tot_conv)) +
  geom_line() + 
  theme_minimal() + 
  ggtitle(" US Conversions by Day") + 
  xlab("Date") + 
  ylab("Total Conversions") + 
  scale_x_date(NULL, date_breaks = "1 month", date_labels = "%b")


```

This plot contains a lot of useful information. To gain insight into how conversions depend on marketing channel, we can use facets. Facets are subplots that display a time series for each marketing channel.


```{r}
conversions %>% 
  filter(country_code == "us") %>% 
  ggplot(aes(x = datestamp, y = conversions)) + 
  geom_line() + 
  facet_wrap(~ marketing_channel) + theme_minimal()
```

Display ads and search engine ads are the dominant marketing channels. Both have a regular pattern that is likely a function of the day of week, with a higher number of conversions during weekdays as compared with weekends. We can explore this feature by aggregating over each weekday and visualizing how the distribution of conversions changes by day.

```{r}
conversions %>% 
  filter(country_code == "us") %>% 
  drop_na() %>% 
  group_by(datestamp) %>% summarise(conversions = sum(conversions)) %>% 
  mutate(day = wday(datestamp, label = T)) %>% 
  ggplot(aes(x = as.factor(day), y = conversions)) + 
  geom_boxplot() + 
  stat_summary(fun = "mean", geom = "point", color = "red") +
  theme_minimal() + 
  ggtitle("Conversion Distribution by Day") + 
  xlab("Day of Week") + 
  ylab("Conversions")

```

Clearly, there are significant changes in the mean level of conversions across the week. This is a form of seasonality. It may be useful to see what the data look like when this weekday effect is removed. To do so, we could visualize the residuals from the following linear regression model:

\begin{align}
\hat{\text{conversions}} = \hat{\beta}_0 + \sum_{j=2}^7 \bigg( \hat{\beta}_j \times 1(\text{weekday = j}) \bigg),
\end{align}

where $j$ indexes the day of week. The residuals from this model consist of each observation minus the mean for that particular weekday. 

```{r}
# EDA via models

mod_df = conversions %>% 
  filter(country_code == "us") %>% 
  drop_na() %>% 
  group_by(datestamp, country_code) %>% 
  summarise(tot_conv = sum(conversions), .groups = "drop") %>% 
  mutate(wday = factor(wday(datestamp, label= T), ordered = F))

mod = lm(tot_conv ~ wday, data = mod_df)

mod_df$resids = residuals(mod)

mod_df %>% 
  ggplot(aes(x = datestamp, y = resids)) + 
  geom_line() + 
  theme_minimal() + 
  ggtitle("US Conversions excluding Weekly Seasonality")
```

This allows us to more clearly see the trend across the date range, removing the effect of the weekly pattern. 

## Multiple Time Series

Often we will want to develop insight into the relationship between several variables. To illustrate, we will use quarterly data on GDP per capita and the Case Shiller Home Price Index (both from the FRED database). 

```{r}
fred = read.csv("https://raw.githubusercontent.com/dbreynol/DS809/main/data/fred_dat.csv")[,-1]
fred$date = ymd(fred$date)

fred %>% 
  mutate(gdp_s = (gdp-mean(gdp))/sd(gdp), shiller_s = (shiller - mean(shiller))/sd(shiller))  %>%
  pivot_longer(-date) %>% 
  filter(name %in% c("gdp_s", "shiller_s")) %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_line(aes(color = name)) + theme_minimal() + 
  ggtitle("Standardized Case Shiller Index versus GPD per Capita")

```

It looks like these two time series track pretty closely to one another. We could fit a linear regression to this data in order to estimate the expected change in the Case Shiller Index for a unit ($1) change in GDP/ capital

```{r}
m0 = lm(shiller ~ gdp, data = fred)
m0 %>% tidy() %>% kable()
```

Further, we could also examine the residuals to gain insight into what is missing from this model.

```{r}
fred$res = residuals(m0)
ggplot(fred, aes(x = date, y = res)) + geom_line() + theme_minimal() + 
  ggtitle("Residuals from Shiller ~ GDP/Capita linear regression")
```

The model severely underestimates the house index starting during the pandemic. There is a clear pattern to these residuals. Is this a problem?

## Autocorrelation

One of the assumptions of the linear regression model is that the errors are independent and identically distributed. That is, for the model,

\begin{align}
y = X \beta + \epsilon,
\end{align}

The error vector, $\epsilon \sim N(0, \sigma^2)$. This implies that there is no correlation structure to the residuals. One way to check that this is true is to check for the absence of autocorrelation in the observed residuals.

Given a time series, $y_t$, where $t=1,\ldots,T$, the correlation between $y_t$ and its lagged value, $y_{t-k}$, is defined as autocorrelation. Let's first review correlation. For two vectors of data, $x$ and $y$, the correlation between the two is,

\begin{align}
\text{cor}(x,y) &= \frac{\text{cov}(x,y)}{\sigma_x \sigma_y} \\
                &= \frac{ \sum_i (x_i - \bar{x})(y_i - \bar{y})}{ \sqrt{ \sum_i (x_i - \bar{x})^2 \sum_i (y_i - \bar{y})^2 }}
\end{align}

```{r echo=TRUE, eval=FALSE}
set.seed(1)
n = 5
x = rnorm(n)
y = rnorm(n)

sum( (x - mean(x)) * (y - mean(y))) / ( (n-1) * sd(x) * sd(y) )
cor(x, y)
```

Autocorrelation applies this concept to a single time series. Autocorrelation is the correlation of a time series with a delayed copy of itself as a function of delay. The autocorrelation for a time series $y$ at lag $k$ is:

\begin{align}
r_k &= \frac{\text{cov}(y_t, y_{t-k})}{\sigma_{y_t} \sigma_{y_{t-k}}} \\
&= \frac{ \sum_{t = k + 1}^{T} (y_t - \bar{y})(y_{t-k} - \bar{y}) }{\sum_{t=1}^T (y_t - \bar{y})^2 }.
\end{align}

The second line is due to the assumption of constant variance. Here is a simple example of computing the lag 1 autocorrelation.

```{r echo=TRUE, eval=FALSE}
a = c(1,2,3,4,5)

a1 = c(1,2,3,4)
a2 = c(2,3,4,5)

# lag 1 autocorrelation
sum( (a1 - mean(a)) * (a2 - mean(a))) / (sum( (a - mean(a))^2 ) )  # by hand
(acf(a))
```


## Lab 1

1. Starting from the code chunk in [Time Series Data], extend the simulated training data to a full week (December 11 through December 15, eight hours each day). Using the data frame representation, <code>plot(trading_dat\$hr, trading_dat\$pr)</code>. Using the time series data, <code>plot(trading_ts)</code>. What are the differences between these two plots?

2. We can further hone in on the trend of the residuals at the end of [Time Series EDA] by computing and plotting a moving average. For a time series $y_t$, $t = 1, \ldots, T$, a moving average of order $m$ can be written,
\begin{align}
\hat{y_t} = \frac{1}{m} \sum_{j=-k}^{k} y_{t+j},
\end{align}
where $m=2k+1$. The concept behind this technique is that observations that are close in time are likely to be close in value. Compute a moving average of order $m=7$ for the residual time series and plot it along with the residuals in a single plot. 

3. Join the Kayak visits data (see below) to the conversions data from [Time Series EDA]. Use the <code>inner_join</code> function with the argument, <code>by = c("datestamp", "country_code", "marketing_channel")</code>. Make a plot of <code>user_visits</code> and <code>conversions</code> by day. Standardize them if it makes sense. Then, fit a linear regression model with <code>conversions</code> as the reponse variable and <code>user_visits</code> as the explanatory variable. What is the estimated line equation? Finally, make a plot of the residuals from this model.

```{r echo=TRUE, eval=FALSE}
visits = read.csv("https://raw.githubusercontent.com/dbreynol/DS809/main/data/visits.csv")
visits$datestamp = ymd(visits$datestamp)
```

4. Wrtie a function in R that takes two arguments: a time series ($y$) formatted as a vector and an integer ($k$) specifying a lag. The output for this function is the lag $k$ autocorrelation for $y$, using the formula in [Autocorrelation]. Compare the output of your function to the output from <code>(acf(y))</code>. An extra challenge is to allow $k$ to be a vector of lags, in which case your function should return a vector of autocorrelation values.
