---
title: "Lecture 2"
output: ioslides_presentation
date: "2025-01-26"
---

## Stationarity & the mean

Consider a single value, $x_t$

$\text{E}(x_t)$ is taken across an ensemble of _all_ possible time series


## Stationarity & the mean

```{r station_in_mean}
nn <- 200
tt <- 40
ww <- matrix(rnorm(nn*tt), tt, nn)

mm <- apply(ww, 1, mean)

par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))

matplot(ww, type="l", lty="solid",  las = 1,
        ylab = expression(italic(x[t])), xlab = "Time",
        col = gray(0.5, 0.2))
points(rep(0,tt), pch = "-", col = "blue", cex=1.5)
```


## Stationarity & the mean

```{r ex_ts_plot_joint_dist, fig.cap="Our single realization is our estimate!"}

par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))

matplot(ww, type="l", lty="solid",  las = 1,
        ylab = expression(italic(x[t])), xlab = "Time",
        col = gray(0.5, 0.2))
lines(ww[,1], col = "blue", lwd = 2)
```


## Stationarity & the mean

If $\text{E}(x_t)$ is constant across time, we say the time series is _stationary_ in the mean


## Stationarity of time series

_Stationarity_ is a convenient assumption that allows us to describe the statistical properties of a time series.

In general, a time series is said to be stationary if there is

1. no systematic change in the mean or variance  
2. no systematic trend  
3. no periodic variations or seasonality

